### 2012-05-10 Rctl/inst/doc/manual.Rnw
[Rctl] More manual updates

{\color{blue} \tt \bfseries University of Groningen \\
Groningen Bioinformatics centre \& GCC \\
revision \# 1 \\[24pt]
XX Okt 2011}

@@

\textbf{\sffamily Abstract}: Correlated Trait Loci (CTL) mapping using the R package 'ctl'
\vfill

\noindent \emph{email}: \verb|Danny.Arends@Gmail.com|
\newpage

@@

\newpage

@@

After you have stored the data as plain text open the files in a text editor (NOT WORD, for 
windows Notepad++ is a good option) and check to see if your data is stored correctly, if 
not reformat your data in both the genotype.csv and the phenotype.csv files to match the 
format below. {\color{red} \emph{\bf{NOTE: Individuals must match between the genotype and 
the phenotype file...}}} The phenotypes.csv file containing the genotype matrix is stored 
individuals x genetic marker:

@@

\emph{Use an R/qtl formatted dataset}\\
Provided is the main interface functions to R/qtl: $CTLscan.cross()$ this functions accepts and 
R/qtl formatted cross object as input and will scan CTL, perform permutations and transform the
detected differential correlation to LOD score matrices.
As an example we show the code to scan the multitrait dataset provided in R/qtl:

<<echo=TRUE,eval=FALSE>>=
require(qtl)        #Loads the R/qtl package
data(multitrait)    #Loads the dataset
multitrait          #Print basic dataset information
@

\emph{Adding genetic map information}\\
This step is optional, but plots and generated networks look much nicer when chromosome locations
of genetic markers are supplied. The 'mapinfo' object is matrix with 3 columns: "Chr" - the 
chromosome number, "cM" - the location of the marker in centiMorgans and the 3rd column "Mbp" - The 
location of the marker in Mega basepairs. The structure of the $mapinfo.csv$ file:

@@

<<echo=TRUE,eval=FALSE>>=
mapinfo <- read.csv("mapinfo.csv",row.names=1,col.names=TRUE)
traits[1:5,1:3]     #Show the first 5 marker records
@
Finally... after all this we can start scanning QTL and CTL...
\bigskip

@@

\emph{CTLscan options}\\
The genotype matrix and the $geno.enc$ parameter, the phenotype matrix, the important $n.perms$ 
parameter, which should be set to as many as possible, especially when publishing any results, 
but for a quick scan 100 is ussualy sufficient.

@@

@@

Big datsets require special care, R isn't the best platform to deal with large scale genomics data 
from micro- or tilling arrays (Just to mention RNAseq and full DNA strand sequencing). Thus we also
provide a commandline version of out CTL mapping tool: mapctl.exe, for more information on all the 
command line parameters can be found at http://www.mapctl.org/mapctl.html
The mapctl commandline tool uses the same input files as the R version, and supports qtab the new 
genotype and phenotype encoding scheme used by qtlHD.
It provides only scanning and permutation features, the output is analyzed in R as outlined above. 
To load the results from a mapctl scan into R by specifying the input files and the output directory 
created by the tool using:

@@

>>>>

{\color{blue} \tt \bfseries University of Groningen \\
Groningen Bioinformatics Centre \& GCC \\
Revision \# 1 \\[24pt]
XX Okt 2011}

@@

\textbf{\sffamily Abstract}: Correlated Trait Loci (CTL) mapping using the R package 'ctl'
\vfill
\noindent \verb|Software:      http://www.mapctl.org/download.html|\\
\noindent \verb|Manual:        http://www.mapctl.org/manual.html|\\
\noindent \verb|Source-code:   http://www.github.com/DannyArends/ctl|\\
\noindent \verb|E-mail:        Danny.Arends@Gmail.com|
\newpage

@@

@@

After you have stored the data as plain text open the files in a text editor (NOT WORD, for 
windows Notepad++ is a good option) and check to see if your data is stored correctly, if 
not reformat your data in both the genotype.csv and the phenotype.csv files to match the 
format below.
{\color{red} \emph{\bf{ \\\\NOTE: Individual order must match between the genotype and the phenotype file... \\\\}}} 
The phenotypes.csv file containing the genotype matrix is stored individuals x genetic marker:

@@

\emph{Use an R/qtl formatted dataset}\\
Provided is the main interface functions to R/qtl: $CTLscan.cross()$ this functions accepts and 
R/qtl formatted cross object as input and will scan CTL, perform permutations and transform the
detected differential correlation to LOD score matrices.
As an example we show the code to load the internal multitrait dataset provided in R/qtl, load
your own by using the $read.cross()$ function:

<<echo=TRUE,eval=FALSE>>=
require(qtl)        #Loads the R/qtl package
data(multitrait)    #Loads the dataset
multitrait          #Print basic dataset information
?read.cross         #List of formats supported by R/qtl
@

\emph{Adding genetic map information}\\
This step is optional, but plots and generated networks look much nicer when chromosome locations
of genetic markers are supplied. The 'mapinfo' object is matrix with 3 columns: "Chr" - the 
chromosome number, "cM" - the location of the marker in centiMorgans and the 3rd column "Mbp" - The 
location of the marker in Mega basepairs.
{\color{red} \emph{\bf{ \\\\NOTE: Marker order must match with the genotype file... \\\\}}} 
The structure of the $mapinfo.csv$ file:

@@

<<echo=TRUE,eval=FALSE>>=
mapinfo <- read.csv("mapinfo.csv",row.names=1,col.names=TRUE)
mapinfo[1:5,1:3]     #Show the first 5 marker records
@
Finally... after all this we can start scanning QTL and CTL...
\bigskip

@@

\emph{CTLscan options}\\
The genotype matrix and the $geno.enc$ parameter, the phenotype matrix, the important $n.perms$ 
parameter, which should be set to as many as possible, especially when publishing any results, 
but for a quick scan 100 is ussualy sufficient:
<<echo=TRUE,eval=FALSE>>=
ctls_quick_scan      <- CTLscan(geno,traits,n.perm=100)
ctls_for_publication <- CTLscan(geno,traits,n.perm=15000)
@

@@

\emph{CTLnetwork and cytoscape}\\
The CTLnetwork function outputs CTL and (optional) QTL data from the CTLobject into a network 
format, if mapinfo is available it is used. The network format is relatively easy and can also
be parsed by many other programs. As an example we use cytoscape, but first lets generate some 
network input files:

<<echo=TRUE,eval=FALSE>>=
ctl_network <- CTLnetwork(ctls, lod.threshold = 5, mapinfo)
@

The CTL edges in the network\_full.sif file are annotated as:

\begin{tabular}{lrrrrr}
Trait & CTL & Trait & LOD Score\\
\end{tabular}

The CTL edges in the network\_summary.sif file look like this:

\begin{tabular}{lrrrrr}
Trait & CTL & Trait & avgLOD Score & sumLOD Score & nTraits\\
\end{tabular}

Additional QTLs are added to both files in the folowing structure:

\begin{tabular}{lrrr}
Trait & QTL & Marker & LOD Score\\
\end{tabular}

This allows the first additional infromation column (Nr 4) to act as an distance measurement, 
the closer two traits or two markers are together in the network the stronger a detected QTL 
or a CTL.

@@

Big datsets require special care, R isn't the best platform to deal with large scale genomics data 
from micro- or tilling arrays (Just to mention RNAseq and full DNA strand sequencing). Thus we also
provide a commandline version of our CTL mapping routine: mapctl.exe, for more information on all 
the command line parameters can be found at http://www.mapctl.org/mapctl.html
The mapctl commandline tool uses the same input files as the R version, and supports qtab the new 
genotype and phenotype encoding scheme used by qtlHD. It provides only QTL and CTL scanning and 
permutation features, the output is analyzed in R as outlined above. To load the results from a 
mapctl scan into R by specifying the input files and the output directory created by the tool 
using:

@@

\bigskip
\textbf{\sffamily Some famous last words}
\nopagebreak
